\myTitle

### 2015-03-12 tex/response_to_refs_letter.tex

\begin{letter}{ \today

Dr. Nolan Kane\\
News and Views Editor\\
Molecular Ecology Resources \\
}

\opening{Dear Dr.~Kane:}

Thank you for the opportunity to resubmit our revised manuscript, MER-14-0343 ``Interpreting the FLOCK algorithm
from a statistical perspective.''  We are very grateful to you and to both of the referees for the many
suggestions that we have incorporated.  We feel they have improved the manuscript greatly.  We have taken your
advice and undertaken a set of simulations that provide a thorough assessment of {\sc flock} versus {\sc 
structure} on data sets with large numbers (2,000) of individuals, like our original steelhead data set.
We have dropped the steelhead data set from the paper.

In order to conduct these simulations we implemented the {\sc flock} algorithm by modifying the source code of
the program {\sc structure}.  We used the resulting program, {\sc flockture}, because {\sc flock} itself
is 200 times slower than {\sc flockture} and it is also not amenable to scripting and repeated application to 
multiple data sets in batch mode.  It is simply not possible to prepare a reproducible set of simulations using
the point-and-click interface of {\sc flock}.  We show simulations in the paper to verify that {\sc flockture} and {\sc flock} yield equivalent results before comparing {\sc structure} to {\sc flockture}.

Another change in the paper is that we focus on the performance of the programs for clustering individuals while
assuming the correct value of $K$.  We were actually somewhat uncomfortable in our initial submission that
assessing {\sc flock}'s ability to estimate $K$ might give the impression that we recommend such pursuits---an
endeavor that I (along with {\sc structure} co-author Matthew Stephens) are reluctant to recommend, especially
when population structure is weak.  So, we are pleased to be able to offer a rigorous evaluation of program
performance without going too deeply into the {\em ad-hoc} procedures for estimating $K$ with {\sc flock}.

Overall, we discovered through our simulations that results from the {\sc flock} algorithm tend to be less 
consistent than {\sc structure}'s, especially at low values of $F_\mathrm{ST}$.  But, on the other hand, we were 
quite interested to see that the algorithm still worked quite well in some cases.  In fact, it outperformed {\sc structure} 
using SNPs at intermediate levels of differention.  We also found that in large data sets like those that
we simulated, {\sc flock}'s individual runs seldom (or never) converge to exactly the same partition.  This 
presents some serious problems for applying the Duchesne \& Turgeon (2012) recipe for estimating $K$, and we 
note that in the discussion.

Below, we address individual comments from you and the referees.  Our replies are in \textcolor{blue}{{\sl blue slant}} font.

Thank you again for your service.  As usual, we have been greatly pleased with the review process
at {\em Molecular Ecology Resources} and are excited with how much improvement the referees have brought
to our paper.  We apologize that it has required several months to return these revisions, but do appreciate
the extra time the journal granted us---it was necessary to write {\sc flockture} and to undertake the
simulations.

Best regards,

Eric C. Anderson

\vspace*{3em}

@@

This paper, once revised, has the potential to shed light on some of the key similarities and differences among STRUCTURE and FLOCK, both of which are useful and important programs for population genetic analysis.  In that, I agree heartily with reviewer 2. However, reviewer 1 raises many important points that must be dealt with if this is to become acceptable. In particular, it would be important to broaden the scope of the article beyond a single dataset.  There are numerous well-analyzed datasets available these days.  For instance, take:

Recommendations for utilizing and reporting population genetic analyses: the reproducibility of genetic clustering using the program STRUCTURE
KJ Gilbert, RL Andrew, DG Bock, ...
Molecular Ecology 21 (20), 4925-4930

for some approaches that we used to assess repeatability of analytical procedures in quite a number of Structure datasets.  It would not be necessary to go quite as in-depth as in that article, but I do agree with reviewer 1 that a single dataset is insufficient.  Additionally it would be very helpful to use either simulated datasets or other datasets where we have a good idea of what the 'true' answer should be (e.g. some independent prior knowledge of some kind, or at the least a very thorough existing analysis to give reasonable expectations).

\reply{Thank you for your recommendations.  We have undertaken a set of simulations as suggested.}

@@

Reviewer 1 raises numerous other important points which will improve this paper substantially when dealt with, as does Reviewer 2 (though this reviewer's concerns tend to be much more minor). Despite the large number of suggestions between the reviewers, is should be quite possible to make these changes relatively easily - with the availability of so many existing, well-analyzed datasets it should be relatively easy to run at least a few more through FLOCK and STRUCTURE for a broader set of comparisons.  Alternately, it might make sense to do a fuller analysis of many datasets and re-submit as a full paper. Either would be quite useful to the Molecular Ecology and MER readership.

\reply{We address all reviewer comments in the revision, and detail our responses below.  We are re-submitting
our manuscript as a revision of our first version; however if you thought that it warranted
reclassification and publication as
a full paper rather than as a comment we would be amenable to that.  Thank you.}

@@

I now describe each of those concerns in more detail.

{\bf 1) Conceptual framework}

It must be emphasized that FLOCK is not an MCMC based algorithm. It has no target function, no transition probabilities and does not move in small measured probabilistic steps. No amount of mathematical analysis will ever change this.

\reply{We didn't say {\sc flock} was an MCMC algorithm.  We noted that it was a limiting
 case of a simulated annealing algorithm.}

In fact, similar iterative methods have been around for a very long time (e.g. Newton's method for finding roots), much before the first MCMC algorithms were designed (1950?s). This family of methods should be analysed within a truly pertinent conceptual framework (attractors, basins of attraction, fixed points, orbits, etc). The book A First Course in Discrete Dynamical Systems (RA Holmgren, 1994) provides an excellent introduction to this topic. 
\reply{The referee is correct that iterative methods have been around for a long time.
In fact, rather than viewing it as special case of simulated annealing, one could just as well
note that {\sc flock} is a type of gradient descent-algorithm attempting to maximize the
likelihood of the data given a configuration of clusters.
Regardless of what name is given to it, since it is effectively an algorithm
that (intentionally or not) yields solutions with 
high (and possibly maximal) likelihood, we find it quite clear that  
the conceptual framework of statistics is at least as ``truly pertinent,''
for purposes of analysis of the algorithm, as is
the conceptual framework of dynamical systems.
}

@@

The current popularity of MCMC methods in the field of genetic analysis does not justify trying to cast every current algorithm into an MCMC framework.

\reply{Again, we are not so much casting it in an MCMC framework as we are in a 
statistical/probabilistic framework.  Since it is a problem in statistical inference
it seems to us reasonable to understand it in terms of the probabilities involved. Those
probabilities are, for the most part, identical to those in the program {\sc structure} because
the underlying model is the same in the two programs.  Regarding casting the algorithm in
comparison to {\sc structure}, it makes sense to us to try to understand and explain the {\sc flock}
algorithm in those terms since {\sc structure} is so well known and because the models
underlying it and {\sc flock} are essentially identical.  }

For example:
\begin{itemize}
\item Line 156: it is suggested that Flock can be trapped in a local mode, this is incorrect because Flock does NOT search for maximum values, either local or else
\reply{We thank the referee for pointing out our erroneous language here.  We should not have used
the word ``trapped.''  We have updated that sentence to read, ``As such, we should not be surprised
if it is prone 
to converging to different parts of the posterior probability surface on runs started
from different initial conditions.''}

\item Line 254: it is suggested that Flock is searching over all possible partitions. Flock does not sample nor search the surface or space of all partitions.
\reply{We would not have wanted it to sound like we thought {\sc flock} was searching
each and every possible
partition, since this would be infeasible in most problems, so we thank the referee for pointing out that our reference to a space of all partitions might
have confused readers in this way.  We have changed that section appropriately.  However, on whether
or not {\sc flock} samples from the space of partitions, we contend that the referee is
incorrect when he states that, ``Flock does not sample nor search the surface or space of all partitions.'' 
Consider this: 1) for each run, the initial starting partition in {\sc flock} is chosen (or you could say,
sampled) randomly from amongst
possible partitions with roughly the same number of individuals in each cluster; 2) the starting partition uniquely
determines the partition that the algorithm will converge to since it is deterministic, 3) different
starting partitions can lead to different final solutions/partitions.  How is it then that each
run in {\sc flock} is {\em not} a sample from the partitions? It is quite clear to us that it is.  }
\item Line 234: ``individuals would alternate between reference groups'': this, precisely, is an orbit
\reply{We removed discussions of individuals alternating between reference groups.}
\end{itemize}

@@

{\em Misinterpretation of `best run'}

However, the authors are in fact proposing to analyse the output of FLOCK for each K as if it were potentially informative. This, in turn, lead to a misunderstanding of what `best run' means. For instance on p.12, they state:
``For all groups of runs for $K> 2$ no plateaus were observed...it is unclear in these cases what constitutes the `best run' as all plateau lengths are equal to one.''

The answer is simple: there is no best run. The term `best run' applies exclusively to the run with the longest plateau corresponding to the estimate of K. First the stopping rules are applied then the estimation rules. If the estimate is not undecided, the only best run to be considered is the one associated with the estimated K. If the estimate is undecided there is no best run. The documentation and also the flow chart in the latest article (Duchesne and Turgeon 2012) make that clear. In fact, outside the estimated K, no partitions from FLOCK should be considered. Therefore comparisons between clusterings from FLOCK and STRUCTURE for all K except the estimated one are just irrelevant and not valid.

\reply{This seems strange to us.  What happens in the situation where the
``true $K$'' is known and you just want to cluster individuals? If there are no plateaus
amongst the runs at that value of $K$ should one discard {\sc flock} and use
a different program altogether? In most of the simulated data sets we investigated {\sc flock},
like {\sc flockture},
would not have seen any plateaus at all!---and yet it is still delivering useful clustering
inference that is competitive with, and at times better than, {\sc structure}.  We discuss
this rather restrictive guidance on using {\sc flock} in the Discussion.  }

@@

This same approach (= all K outputs are valuable) also brings the authors to lament ``FLOCK's tendency to converge to different solutions with large K'' and conclude that STRUCTURE would be a better choice for complex problems. This is a surprising conclusion since FLOCK is expected to signal the absence of K components precisely by not repeating solutions for that K and thus producing very short plateaus if any. Clearly, by increasing the value of K, the true value will eventually be surpassed and FLOCK will tend to generate numerous solutions and therefore short plateaus ($< 6$). As for the real dataset, there is no ground to believe that $K > 4$ (line 246--248), so that the reported large variability of the FLOCK results may just be the right signal. Here there is a huge misunderstanding that leads the authors to systematically misinterpret potentially correct answers as a sign of inconsistency or lack of power.

\reply{These points are now largely moot, as we don't focus on the problem of estimating $K$, and we
only analyze simulated data sets at the correct $K$ value.}

@@

\reply{We agree with the referee that the present version of our manuscript should be very useful.}

They should be published in a full-fledged article, not just a comment.

\reply{If, upon reading the present version of our manuscript, the editor feels it would be appropriate
to publish it as a full article, we would agree to that.
}

I also suggest that the authors make an additional effort to better understand plateau analysis especially in connection with variability of solutions i.e. lengths of plateaus. Short plateaus (numerous solutions) are the right answer whenever K is the wrong answer to the `number of populations' problem.

\reply{We have taken the referee's suggestion and worked on obtaining a better understanding of plateau analysis.
Short plateaus may well be right answer whenever $K$ is the wrong answer.  However, short plateaus are also
the wrong answer when the $K$ value is correct.  This seems to be major problem for {\sc flock} on data sets
of the size that we have simulated.  We've addessed this in the 2nd to last paragraph of the discussion. }

If run times are to be compared, the comparisons should follow the above recommendation such that a real user gets a simple message out of it.

\reply{See above.}
Pierre Duchesne

\end{letter}
